---
title: "Getting Started with birt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with birt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is birt?

**birt** (Bayesian IRT) is an R package for fitting Item Response Theory models
using Bayesian estimation via CmdStan. It supports three dichotomous IRT models:

- **Rasch (1PL)**: Every item has the same discrimination — items differ only in difficulty.
- **2PL**: Each item gets its own discrimination parameter controlling how sharply
  it separates students of different ability.
- **3PL**: Adds a guessing parameter representing the probability of a correct
  response by pure chance.

All models are estimated using Hamiltonian Monte Carlo via CmdStan, producing full
posterior distributions for every parameter. This means you get credible intervals,
convergence diagnostics, and posterior predictive checks — not just point estimates.


## Installation

birt requires CmdStan installed on your computer.

```{r}
# Step 1: Install cmdstanr (the R interface to CmdStan)
install.packages("cmdstanr", repos = c(
  "https://stan-dev.r-universe.dev/",
  getOption("repos")
))

# Step 2: Install CmdStan itself (~5 minutes)
cmdstanr::install_cmdstan()

# Step 3: Install birt from GitHub
# install.packages("remotes")
remotes::install_github("Ndukaboika/birt")
```


Verify the installation:

```{r}
cmdstanr::cmdstan_version()
library(birt)
```

## The Models

All three models share a core idea: the probability of a correct response depends on
the difference between a person's ability and an item's difficulty. They differ in how
many item parameters they estimate.

### Rasch Model (1PL)

$$\text{logit}(P(Y_{jk} = 1)) = \alpha_j + \delta - \beta_k$$

Where:

- $\delta$ = overall mean ability
- $\alpha_j$ = how student $j$ deviates from the mean
- $\beta_k$ = difficulty of item $k$
- Total ability: $\theta_j = \alpha_j + \delta$

When ability equals difficulty ($\theta_j = \beta_k$), the student has a 50% chance
of answering correctly.

### 2PL Model

$$\text{logit}(P(Y_{jk} = 1)) = a_k \cdot (\alpha_j + \delta - \beta_k)$$

Adds discrimination $a_k$. When $a_k > 1$, the item separates students more sharply.
When $a_k < 1$, the item is less discriminating. When $a_k = 1$, this reduces to the
Rasch model.

### 3PL Model

$$P(Y_{jk} = 1) = c_k + (1 - c_k) \cdot \text{logistic}(a_k \cdot (\alpha_j + \delta - \beta_k))$$

Adds guessing $c_k$. Even very low-ability students have probability $c_k$ of getting
the item right. For 4-option multiple choice, expect $c_k \approx 0.25$.

### Model Comparison

| Feature | Rasch | 2PL | 3PL |
|---------|-------|-----|-----|
| Difficulty ($\beta$) | Yes | Yes | Yes |
| Discrimination ($a$) | Fixed at 1 | Estimated | Estimated |
| Guessing ($c$) | Fixed at 0 | Fixed at 0 | Estimated |
| Parameters per item | 1 | 2 | 3 |
| Minimum sample size | ~100 | ~200 | ~500 |

## Prior Distributions

birt uses weakly informative default priors that work across a wide range of testing
scenarios. The defaults are deliberately neutral — they make no strong assumptions
about your test or students. Users with domain knowledge can override any prior at
two levels: class-level (same prior for all items) or per-item (different prior for
each individual item).

### Default Priors

| Parameter | Default Prior | Class-level Argument | Per-item Arguments | 95% Range |
|-----------|-------------|---------------------|-------------------|-----------|
| $\delta$ (mean ability) | Normal(0, 1) | `prior_delta = c(0, 1)` | — | -2.0 to 2.0 |
| $\alpha_j$ (ability deviation) | Normal(0, 1.5) | `prior_alpha_sd = 1.5` | — | -3.0 to 3.0 |
| $\beta_k$ (difficulty) | Normal(0, 1.5) | `prior_beta = c(0, 1.5)` | `prior_beta_mean`, `prior_beta_sd` | -3.0 to 3.0 |
| $a_k$ (discrimination) | LogNormal(0, 0.5) | `prior_a = c(0, 0.5)` | `prior_a_meanlog`, `prior_a_sdlog` | 0.37 to 2.72 |
| $c_k$ (guessing) | Beta(2, 8) | `prior_c = c(2, 8)` | `prior_c_alpha`, `prior_c_beta` | 0.03 to 0.45 |

These defaults are informed by recommendations from the Stan User's Guide (Stan
Development Team, 2024), Luo and Jiao (2018), and the edstan package (Furr, 2017).
The ability prior also serves to identify the scale of the model.

### Why These Defaults?

- **delta ~ Normal(0, 1)**: Centered at zero — no assumption about whether students
  are above or below average difficulty. The data determines this.
- **alpha ~ Normal(0, 1.5)**: Wide enough to accommodate very strong and very weak
  students. A deviation of 3 logits shifts P(correct) from 50% to about 95%.
- **beta ~ Normal(0, 1.5)**: Same rationale. Covers the typical difficulty range of
  well-constructed test items.
- **a ~ LogNormal(0, 0.5)**: Always positive (negative discrimination contradicts IRT
  assumptions), centered at 1.0 (the Rasch case).
- **c ~ Beta(2, 8)**: Weakly informative with a mean around 0.2. Wide enough to
  accommodate different item formats without making strong assumptions.

### Customizing Priors: Class-Level

Set the same prior for all items of a parameter type:

```{r}
# If you know students tend to perform above average
fit <- rasch_fit(data,
  prior_delta = c(0.75, 0.5),
  seed = 123
)

# Wider difficulty prior for all items
fit <- rasch_fit(data,
  prior_beta = c(0, 3),
  seed = 123
)

# 4-option multiple choice (guessing around 0.25)
fit3 <- threepl_fit(data,
  prior_c = c(5, 15),    # Beta(5,15), mean = 0.25
  seed = 123
)

# 5-option multiple choice (guessing around 0.20)
fit3 <- threepl_fit(data,
  prior_c = c(5, 20),    # Beta(5,20), mean = 0.20
  seed = 123
)

# Free-response items (very little guessing)
fit3 <- threepl_fit(data,
  prior_c = c(1, 19),    # Beta(1,19), mean = 0.05
  seed = 123
)

# Wide priors for large samples (let data fully dominate)
fit <- rasch_fit(data,
  prior_delta    = c(0, 3),
  prior_alpha_sd = 3,
  prior_beta     = c(0, 3),
  seed = 123
)
```

### Customizing Priors: Per-Item

Set a different prior for each individual item. Use this when you have specific
knowledge about certain items from pilot testing, expert judgment, or previous
administrations.

```{r}
K <- ncol(data)

# Difficulty: item 3 is known to be hard, item 7 is known to be easy
b_mean <- rep(0, K)        # default for all items
b_mean[3] <- 2.0           # item 3: prior centered at 2.0 (hard)
b_mean[7] <- -1.5          # item 7: prior centered at -1.5 (easy)

b_sd <- rep(1.5, K)        # default uncertainty for all items
b_sd[c(3, 7)] <- 0.5       # tighter prior for items we know about

fit <- rasch_fit(data,
  prior_beta_mean = b_mean,
  prior_beta_sd = b_sd,
  seed = 123
)
```

```{r}
# Discrimination: item 5 is known to be poorly discriminating
a_meanlog <- rep(0, K)       # default: centered at 1.0 for all items
a_meanlog[5] <- -0.5         # item 5: prior centered below 1.0

a_sdlog <- rep(0.5, K)       # default uncertainty for all items
a_sdlog[5] <- 0.3            # tighter prior for item 5

fit2 <- twopl_fit(data,
  prior_a_meanlog = a_meanlog,
  prior_a_sdlog = a_sdlog,
  seed = 123
)
```

```{r}
# Guessing: mixed item formats on the same test
# Items 1-5 are 4-option MC, items 6-10 are 5-option MC
c_alpha <- rep(5, K)
c_beta  <- c(rep(15, 5), rep(20, 5))   # mean 0.25 vs 0.20

fit3 <- threepl_fit(data,
  prior_c_alpha = c_alpha,
  prior_c_beta = c_beta,
  seed = 123
)
```

Per-item arguments override class-level arguments. If you provide
`prior_beta_mean`, the `prior_beta` argument is ignored for the mean.
If you provide `prior_beta_sd`, the `prior_beta` argument is ignored for the SD.

### Checking What Priors Were Used

The priors are stored in the fitted object as vectors (one value per item):

```{r}
fit <- rasch_fit(data, seed = 123)
fit$priors

# Shows:
# $delta     — c(0, 1)
# $alpha_sd  — 1.5
# $beta_mean — c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0)  one per item
# $beta_sd   — c(1.5, 1.5, 1.5, ...)              one per item
```

## Quick Start with Simulated Data

### Simulate Data

`rasch_simulate()` generates data with known true parameters so you can verify
the model recovers the correct values.

```{r}
library(birt)

sim <- rasch_simulate(
  J = 300,            # 300 students
  K = 10,             # 10 items
  delta_true = 0.75,  # mean ability
  alpha_sd = 1,       # spread of abilities
  seed = 42           # reproducible
)

# The response matrix
head(sim$data)

# True values we want to recover
sim$beta    # item difficulties
sim$delta   # mean ability
```

### Fit the Rasch Model

```{r}
fit <- rasch_fit(
  data = sim$data,
  chains = 4,
  parallel_chains = 4,
  iter_sampling = 1000,
  seed = 123
)

# Quick overview
fit

# Detailed summary
summary(fit)
```

### Understanding the Output

The `summary()` output contains:

**Mean Ability (delta)**: The estimated overall mean ability with a credible interval.
Compare this to `sim$delta` to check recovery.

**Item Difficulties (beta)**: One row per item with the posterior mean (point estimate),
credible interval, and convergence diagnostics. Check that Rhat < 1.01 and ESS > 400
for all items.

**Person Ability Summary**: Overview of estimated abilities across all students.

### Extract Parameters

```{r}
# Item difficulties
item_params(fit)

# Person abilities (total: alpha + delta)
head(person_params(fit), 10)

# Mean ability
delta_param(fit)

# Change credible interval width
item_params(fit, prob = 0.90)  # 90% interval
item_params(fit, prob = 0.99)  # 99% interval
```

### Item Fit Diagnostics

Outfit and infit are mean-square statistics measuring how well each item
conforms to the model. Both should be near 1.0.

```{r}
ifit <- item_fit(fit)
ifit

# Flag misfitting items
ifit[ifit$outfit > 1.3 | ifit$outfit < 0.7, ]
```

Interpretation:

- **Outfit > 1.3**: The item behaves erratically — unexpected responses from students
  far from the item's difficulty level.
- **Outfit < 0.7**: The item is too predictable — possibly redundant with other items.
- **Infit**: Same interpretation but more sensitive to systematic patterns near the
  item's difficulty and less affected by outliers.

### Person Fit Diagnostics

```{r}
pfit <- person_fit(fit)
head(pfit)

# Unusual response patterns
pfit[pfit$outfit > 1.3, ]
```

High person outfit may indicate guessing, careless responding, or cheating.

### Plots

#### Item Characteristic Curves

Each curve shows P(correct) as a function of ability. Harder items are shifted right.
Shaded bands show 95% credible intervals.

```{r}
plot(fit, type = "icc")

# Specific items only
plot_icc(fit, items = c(1, 5, 10))

# Without credible bands
plot_icc(fit, ci = FALSE)
```

#### Wright Map

Places students (histogram) and items (triangles) on the same logit scale. This
reveals whether the test is well-targeted:

- Items covering the full student range = good targeting.
- All items to the left of students = test is too easy.
- Gaps with no items = poor measurement in that ability range.

```{r}
plot(fit, type = "wright")
```

#### Test Information Function

Shows where on the ability scale the test is most precise. Higher information
means more precise measurement.

```{r}
# Total test information
plot(fit, type = "info")

# With individual item curves
plot_info(fit, show_items = TRUE)
```

#### Trace Plots

Check MCMC convergence. Chains should overlap and look like "fuzzy caterpillars."
If chains are stuck in different places or trending, the model hasn't converged.

```{r}
plot(fit, type = "trace")
```

### Parameter Recovery

With simulated data, verify the model recovers the truth:

```{r}
items_est <- item_params(fit)

# Plot true vs estimated
plot(sim$beta, items_est$mean,
     xlab = "True Difficulty",
     ylab = "Estimated Difficulty",
     pch = 19, main = "Parameter Recovery")
abline(0, 1, col = "red", lty = 2)

# Correlation (should be > 0.95)
cor(sim$beta, items_est$mean)

# Delta recovery
cat("True delta:", sim$delta, "\n")
cat("Estimated delta:", delta_param(fit)$mean, "\n")
```

## Fitting the 2PL Model

Use the 2PL when you suspect items differ in how well they discriminate between
high- and low-ability students.

```{r}
fit2 <- twopl_fit(sim$data, seed = 123)
summary(fit2)
```

### Discrimination Parameters

```{r}
discrim_params(fit2)
```

Interpretation:

- $a \approx 1$: Similar to Rasch — average discrimination.
- $a > 1.5$: Highly discriminating — steep ICC, strongly separates students.
- $a < 0.5$: Poorly discriminating — flat ICC. Consider removing the item.

### 2PL Plots

Notice the different slopes in the ICCs — steeper curves correspond to higher
discrimination:

```{r}
plot(fit2, type = "icc")
plot(fit2, type = "info")
```

### Custom Priors for 2PL

```{r}
# Class-level: wider discrimination prior for all items
fit2_wide <- twopl_fit(sim$data,
  prior_a = c(0, 1),
  seed = 123
)

# Per-item: item 5 is known to be poorly discriminating
K <- ncol(sim$data)
a_meanlog <- rep(0, K)
a_meanlog[5] <- -0.5

fit2_item <- twopl_fit(sim$data,
  prior_a_meanlog = a_meanlog,
  seed = 123
)
fit2_item$priors
```

## Fitting the 3PL Model

Use the 3PL for multiple-choice tests where guessing is plausible. Requires
500+ students for stable estimation.

```{r}
sim_large <- rasch_simulate(J = 500, K = 10, seed = 42)

fit3 <- threepl_fit(sim_large$data, seed = 123)
summary(fit3)
```

### Guessing Parameters

```{r}
guessing_params(fit3)
```

Interpretation:

- $c \approx 0.25$: Typical for 4-option multiple choice.
- $c \approx 0$: No guessing (expected for free-response items).
- $c > 0.35$: Unusually high — the item may have poor distractors.

### 3PL with Class-Level Guessing Priors

```{r}
# 4-option multiple choice
fit3_mc4 <- threepl_fit(sim_large$data,
  prior_c = c(5, 15),   # Beta(5,15), mean = 0.25
  seed = 123
)

# 5-option multiple choice
fit3_mc5 <- threepl_fit(sim_large$data,
  prior_c = c(5, 20),   # Beta(5,20), mean = 0.20
  seed = 123
)

# Free-response items
fit3_fr <- threepl_fit(sim_large$data,
  prior_c = c(1, 19),   # Beta(1,19), mean = 0.05
  seed = 123
)
```

### 3PL with Per-Item Guessing Priors

```{r}
# Mixed test: items 1-5 are 4-option MC, items 6-10 are 5-option MC
K <- 10
c_alpha <- rep(5, K)
c_beta  <- c(rep(15, 5), rep(20, 5))   # mean 0.25 vs 0.20

fit3_mixed <- threepl_fit(sim_large$data,
  prior_c_alpha = c_alpha,
  prior_c_beta = c_beta,
  seed = 123
)
fit3_mixed$priors
```

### 3PL Plots

Notice the lower asymptote — P(correct) never reaches zero:

```{r}
plot(fit3, type = "icc")
```

## Working with Real Data

### The Algebra Dataset

birt includes 1,382 students on a 12-item algebra test.

```{r}
data(algebra)
dim(algebra)
colMeans(algebra, na.rm = TRUE)
```

### Fit and Explore

```{r}
fit_alg <- rasch_fit(algebra, seed = 123)
summary(fit_alg)

# Hardest items
items <- item_params(fit_alg)
items[order(items$mean, decreasing = TRUE), ]

# Item fit
item_fit(fit_alg)

# Plots
plot(fit_alg, type = "icc")
plot(fit_alg, type = "wright")
plot(fit_alg, type = "info")
```

### Compare Rasch vs 2PL on Real Data

```{r}
fit_alg_2pl <- twopl_fit(algebra, seed = 123)

# Are discriminations similar or very different?
discrim_params(fit_alg_2pl)

# Compare ICCs
plot(fit_alg, type = "icc")      # Rasch: parallel curves
plot(fit_alg_2pl, type = "icc")  # 2PL: varying slopes
```

### Per-Item Priors on Real Data

```{r}
# Suppose from pilot testing you know items Alg3 and Alg8 are hard
K <- ncol(algebra)
b_mean <- rep(0, K)
b_mean[3] <- 1.5    # Alg3 is hard
b_mean[8] <- 2.0    # Alg8 is very hard

fit_alg_informed <- rasch_fit(algebra,
  prior_beta_mean = b_mean,
  seed = 123
)

# Compare default vs informed estimates
items_default  <- item_params(fit_alg)
items_informed <- item_params(fit_alg_informed)

cbind(
  item = items_default$item,
  default = round(items_default$mean, 2),
  informed = round(items_informed$mean, 2)
)
```

### Using Your Own Data

Your data should be a matrix or data frame:

- **Rows** = students (examinees, respondents)
- **Columns** = items (questions, tasks)
- **Values** = 0 (incorrect) or 1 (correct)
- **NA** = missing (handled automatically)

```{r}
my_data <- read.csv("my_test_data.csv", row.names = 1)
my_data <- as.matrix(my_data)

# Verify format
dim(my_data)
table(my_data, useNA = "ifany")

# Fit
fit_mine <- rasch_fit(my_data, seed = 123)
summary(fit_mine)
```

## Sensitivity Analysis

A sensitivity analysis checks whether your results depend on the prior. Re-fit
the model with different priors and compare. If estimates barely change, your
conclusions are data-driven.

```{r}
# Default weakly informative priors
fit_default <- rasch_fit(algebra, seed = 123)

# Tighter priors
fit_tight <- rasch_fit(algebra,
  prior_alpha_sd = 0.5,
  prior_beta = c(0, 0.5),
  seed = 123
)

# Very wide priors
fit_wide <- rasch_fit(algebra,
  prior_delta    = c(0, 5),
  prior_alpha_sd = 3,
  prior_beta     = c(0, 3),
  seed = 123
)

# Compare item difficulty estimates
items_default <- item_params(fit_default)
items_tight   <- item_params(fit_tight)
items_wide    <- item_params(fit_wide)

# Correlations above 0.99 indicate results are robust to prior choice
cor(items_default$mean, items_tight$mean)
cor(items_default$mean, items_wide$mean)

# Visual comparison
plot(items_default$mean, items_wide$mean,
     xlab = "Default Priors", ylab = "Wide Priors",
     pch = 19, main = "Prior Sensitivity")
abline(0, 1, col = "red", lty = 2)
```

## Advanced Usage

### Accessing the Raw CmdStan Object

```{r}
# The CmdStanMCMC object
class(fit$fit)

# Raw draws for custom analysis
draws <- fit$fit$draws(variables = "beta", format = "draws_matrix")
dim(draws)

# Save a fitted model
fit$fit$save_object("my_fit.rds")
```

### Controlling MCMC Sampling

```{r}
# More iterations for better estimates
fit <- rasch_fit(data, iter_sampling = 2000, seed = 123)

# Fix divergent transitions
fit <- rasch_fit(data, adapt_delta = 0.95, seed = 123)

# Fewer chains for speed (not recommended for final analysis)
fit <- rasch_fit(data, chains = 2, parallel_chains = 2, seed = 123)
```

## Choosing a Model

1. **Start with Rasch.** Simplest and most interpretable. If item fit statistics
   are acceptable (outfit/infit between 0.7 and 1.3), stop here.

2. **Try 2PL** if Rasch fit is poor or you have theoretical reasons to expect
   varying discrimination. Compare ICCs — if slopes differ meaningfully, the
   discrimination parameter is capturing real differences.

3. **Try 3PL** only with multiple-choice items, 500+ students, and clear evidence
   of guessing. If estimated guessing parameters are all near zero, the 2PL is
   sufficient.

## Troubleshooting

### Divergent Transitions

The sampler had trouble exploring the posterior. Try increasing adapt_delta:

```{r}
fit <- rasch_fit(data, adapt_delta = 0.95, seed = 123)
```

Values between 0.9 and 0.99 usually help.

### Low ESS

Not enough effective samples. Try more iterations:

```{r}
fit <- rasch_fit(data, iter_sampling = 2000, seed = 123)
```

### Slow Compilation

The Stan model compiles to C++ on first use (~30-60 seconds). It is cached
afterward, so subsequent calls are fast.

### Corrupt Database Error (R 4.5)

If you see "lazy-load database is corrupt", reinstall with:

```{r}
remove.packages("birt")
devtools::install(args = "--no-byte-compile")
```

## References

- Luo, Y., & Jiao, H. (2018). Using the Stan program for Bayesian item
  response theory. *Educational and Psychological Measurement, 78*(3), 384-408.
- Stan Development Team (2024). *Stan User's Guide*, Section 1.11:
  Item-Response Theory Models.
- Gelman, A., Jakulin, A., Pittau, M. G., & Su, Y. S. (2008). A weakly
  informative default prior distribution for logistic and other regression
  models. *Annals of Applied Statistics, 2*(4), 1360-1383.
- Culpepper, S. A. (2016). Revisiting the 4-parameter item response model:
  Bayesian estimation and application. *Psychometrika, 81*(4), 1142-1163.
- Furr, D. C. (2017). edstan: Stan models for item response theory.
  R package.
