---
title: "Getting Started with birt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with birt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is birt?

**birt** (Bayesian IRT) is an R package for fitting Item Response Theory models using Bayesian estimation via CmdStan. It currently supports three dichotomous IRT models:

-   **Rasch (1PL)**: The simplest IRT model — assumes all items discriminate equally.
-   **2PL**: Adds an item-specific discrimination parameter that controls how sharply each item separates high-ability from low-ability examinees.
-   **3PL**: Adds a guessing (pseudo-chance) parameter, acknowledging that even very low-ability examinees may guess correctly on multiple-choice items.

### Why Bayesian?

Traditional IRT software uses Maximum Likelihood Estimation (MLE). Bayesian estimation via MCMC offers several advantages:

-   **Full uncertainty quantification**: You get credible intervals for every parameter, not just point estimates.
-   **Works with small samples**: MLE can fail or produce extreme estimates with small samples. Priors stabilize Bayesian estimates.
-   **Missing data handled naturally**: The Bayesian framework simply integrates over unknown responses.
-   **Model diagnostics built in**: Posterior predictive checks and leave-one-out cross-validation are straightforward.

## Installation

birt requires CmdStan (the C++ Bayesian engine) installed on your computer.

```{r}
# Step 1: Install cmdstanr (the R interface to CmdStan)
install.packages("cmdstanr", repos = c(
  "https://stan-dev.r-universe.dev/",
  getOption("repos")
))

# Step 2: Install CmdStan itself (takes a few minutes)
cmdstanr::install_cmdstan()

# Step 3: Install birt from GitHub
# install.packages("pak")
pak::pak("Ndukaboika/birt")
```

Verify everything works:

```{r}
cmdstanr::cmdstan_version()  # Should print a version number
library(birt)
```

## The IRT Models

All three models share the same core idea: the probability of a correct response depends on the difference between a person's ability and an item's difficulty. They differ in how many item parameters they estimate.

### Rasch Model (1PL)

The simplest model. Every item has the same discrimination — items only differ in difficulty.

$$\text{logit}(P(Y_{jk} = 1)) = \alpha_j + \delta - \beta_k$$

Where:

-   $\alpha_j$ = how student $j$ deviates from the mean ability
-   $\delta$ = the overall mean ability in the population
-   $\beta_k$ = the difficulty of item $k$

A student's **total ability** is $\theta_j = \alpha_j + \delta$. When ability equals difficulty ($\theta_j = \beta_k$), the student has a 50% chance of answering correctly.

### Two-Parameter Logistic Model (2PL)

Adds a **discrimination** parameter $a_k$ that controls how steeply each item's probability curve rises. Items with high discrimination are better at separating students of different ability levels.

$$\text{logit}(P(Y_{jk} = 1)) = a_k \cdot (\alpha_j + \delta - \beta_k)$$

When $a_k = 1$, this reduces to the Rasch model. When $a_k > 1$, the item is more discriminating (steeper ICC). When $a_k < 1$, the item is less discriminating (flatter ICC).

### Three-Parameter Logistic Model (3PL)

Adds a **guessing** parameter $c_k$ representing the probability of a correct response by pure chance. This is most relevant for multiple-choice items.

$$P(Y_{jk} = 1) = c_k + (1 - c_k) \cdot \text{logistic}(a_k \cdot (\alpha_j + \delta - \beta_k))$$

For a 4-option multiple choice item, you might expect $c_k \approx 0.25$. The 3PL requires large samples (500+ students) because the guessing parameter is difficult to estimate.

### Comparing the Three Models

| Feature                   | Rasch      | 2PL        | 3PL       |
|---------------------------|------------|------------|-----------|
| Item difficulty ($\beta$) | Yes        | Yes        | Yes       |
| Item discrimination ($a$) | Fixed at 1 | Estimated  | Estimated |
| Guessing ($c$)            | Fixed at 0 | Fixed at 0 | Estimated |
| Parameters per item       | 1          | 2          | 3         |
| Minimum sample size       | \~100      | \~200      | \~500     |

### Priors

All models use the following priors:

| Parameter | Prior | Rationale |
|---------------------------|------------------|---------------------------|
| $\alpha_j$ (ability deviation) | Normal(0, 1) | Most students within 3 units of the mean |
| $\beta_k$ (difficulty) | Normal(0, 1) | Most items between -3 and +3 logits |
| $\delta$ (mean ability) | Normal(0.75, 1) | Slight expectation that students are above average difficulty |
| $a_k$ (discrimination) | LogNormal(0, 0.5) | Always positive, centered around 1.0 |
| $c_k$ (guessing) | Beta(5, 23) | Centered around 0.18, appropriate for multiple-choice items |

## Quick Start: Simulated Data

### Simulate Data

The `rasch_simulate()` function generates fake data with known true parameters. This lets you verify the model recovers the correct values.

```{r}
sim <- rasch_simulate(
  J = 300,          # 300 students
  K = 10,           # 10 items
  delta_true = 0.75,# mean ability
  alpha_sd = 1,     # spread of student abilities
  seed = 42         # reproducible
)

# The response matrix: rows = students, columns = items
head(sim$data)

# True item difficulties (what we're trying to recover)
sim$beta

# True mean ability
sim$delta
```

### Fit the Rasch Model

```{r}
fit <- rasch_fit(
  data = sim$data,      # the response matrix
  chains = 4,           # 4 independent MCMC chains
  parallel_chains = 4,  # run all 4 at once
  iter_sampling = 1000, # 1000 draws per chain = 4000 total
  seed = 123            # reproducible
)

# Quick overview
fit

# Detailed summary
summary(fit)
```

### Understanding the Output

The `summary()` output has four sections:

1.  **Mean Ability (delta)**: The estimated overall mean ability with a credible interval.

2.  **Item Difficulties (beta)**: One row per item showing the estimated difficulty, uncertainty (sd), 95% credible interval, and convergence diagnostics (Rhat, ESS).

3.  **Person Ability Summary**: A quick overview of the distribution of estimated abilities across all students.

4.  **Convergence Diagnostics**:

    -   **Rhat** should be below 1.01 for all parameters. Values above 1.01 indicate the chains haven't converged — try running longer.
    -   **ESS (bulk)** should be above 400. This is the effective number of independent samples. Low ESS means the estimates are noisy.

### Extract Parameters

```{r}
# Item difficulties — one row per item
item_params(fit)

# Person abilities — one row per student
# Shows total ability (alpha + delta)
head(person_params(fit), 10)

# Mean ability (delta) — a single number
delta_param(fit)

# Change the credible interval width
item_params(fit, prob = 0.90)  # 90% interval
item_params(fit, prob = 0.99)  # 99% interval
```

### Model Diagnostics

#### Item Fit

Outfit and infit are mean-square statistics that measure how well each item conforms to the model. Both should be near 1.0.

```{r}
ifit <- item_fit(fit)
ifit

# Flag potentially misfitting items
ifit[ifit$outfit > 1.3 | ifit$outfit < 0.7, ]
```

-   **Outfit \> 1.3**: More variation than expected. The item behaves erratically — students who should get it right sometimes get it wrong, and vice versa.
-   **Outfit \< 0.7**: Less variation than expected. The item is too predictable — it might be redundant with other items.
-   **Infit**: Same interpretation, but more sensitive to systematic patterns near the item's difficulty level and less sensitive to outliers.

#### Person Fit

Same idea, but per student. Identifies unusual response patterns.

```{r}
pfit <- person_fit(fit)
head(pfit)

# Flag potentially misfitting students
pfit[pfit$outfit > 1.3, ]
```

High person outfit might indicate:

-   Guessing (getting hard items right by luck)
-   Careless responding (getting easy items wrong)
-   Cheating or answer copying

### Plots

#### Item Characteristic Curves (ICCs)

Each curve shows the probability of a correct response as a function of ability. Harder items have curves shifted to the right. The shaded bands show 95% credible intervals.

```{r}
# All items
plot(fit, type = "icc")

# Specific items only
plot_icc(fit, items = c(1, 5, 10))

# Without credible bands
plot_icc(fit, ci = FALSE)
```

#### Wright Map

The Wright map places students (blue histogram) and items (red triangles) on the same logit scale. This is essential for evaluating test targeting:

-   If items are clustered where students are, the test measures well.
-   If items are all to the left of the students, the test is too easy.
-   If there are gaps with no items, the test doesn't measure well in that ability range.

```{r}
plot(fit, type = "wright")
```

#### Test Information Function

The information function shows where on the ability scale the test is most precise. Higher information means more precise measurement. In the Rasch model, each item contributes maximum information at its difficulty level.

```{r}
# Test information (total across all items)
plot(fit, type = "info")

# With individual item information curves overlaid
plot_info(fit, show_items = TRUE)
```

#### Trace Plots

Trace plots show the MCMC chains' paths through parameter space. Good chains look like "fuzzy caterpillars" — they overlap and mix well. If chains are stuck in different places or trending, the model hasn't converged.

```{r}
plot(fit, type = "trace")
```

### Parameter Recovery

When using simulated data, you can check how well the model recovers the true parameters:

```{r}
# Compare estimated vs. true item difficulties
items_est <- item_params(fit)

plot(sim$beta, items_est$mean,
     xlab = "True Difficulty",
     ylab = "Estimated Difficulty",
     pch = 19,
     main = "Parameter Recovery: Item Difficulties")
abline(0, 1, col = "red", lty = 2)

# Correlation (should be > 0.95 with 300 students)
cor(sim$beta, items_est$mean)

# Compare delta
cat("True delta:", sim$delta, "\n")
cat("Estimated delta:", delta_param(fit)$mean, "\n")

# Compare person abilities
persons_est <- person_params(fit)
true_theta <- sim$alpha + sim$delta
cor(true_theta, persons_est$mean)
```

## Fitting the 2PL Model

The 2PL model allows each item to have its own discrimination. Use it when you suspect items differ in how well they distinguish between high- and low-ability students.

```{r}
fit2 <- twopl_fit(sim$data, seed = 123)
fit2
summary(fit2)
```

### Discrimination Parameters

```{r}
# Extract discrimination estimates
discrim_params(fit2)
```

Interpretation:

-   $a \approx 1$: Similar to Rasch — average discrimination.
-   $a > 1.5$: Highly discriminating — the item sharply separates students.
-   $a < 0.5$: Poorly discriminating — the item barely tells high-ability from low-ability students. Consider removing it.

### 2PL ICCs

Notice how the curves now have different slopes. Steeper curves correspond to higher discrimination.

```{r}
plot(fit2, type = "icc")
```

## Fitting the 3PL Model

The 3PL adds guessing. Use it for multiple-choice tests where low-ability students can guess correctly.

**Important**: The 3PL needs large samples (500+ students). With fewer students, the guessing parameter is poorly identified and you may see convergence issues.

```{r}
# Simulate a larger dataset for the 3PL
sim_large <- rasch_simulate(J = 500, K = 10, seed = 42)

fit3 <- threepl_fit(sim_large$data, seed = 123)
fit3
summary(fit3)
```

### Guessing Parameters

```{r}
guessing_params(fit3)
```

Interpretation:

-   $c \approx 0.25$: Consistent with 4-option multiple choice (1 in 4 chance of guessing correctly).
-   $c \approx 0.0$: No guessing — expected for free-response items.
-   $c > 0.35$: Unusually high guessing — the item may have poor distractors.

### 3PL ICCs

Notice the lower asymptote — the probability never reaches zero, even for very low ability.

```{r}
plot(fit3, type = "icc")
```

## Working with Real Data

### The Algebra Dataset

birt includes an example dataset of 1,382 students on a 12-item algebra test.

```{r}
data(algebra)
dim(algebra)       # 1382 students, 12 items
head(algebra)
colMeans(algebra)  # proportion correct per item
```

### Fit and Explore

```{r}
# Fit a Rasch model
fit_alg <- rasch_fit(algebra, seed = 123)
summary(fit_alg)

# Which items are hardest?
items <- item_params(fit_alg)
items[order(items$mean, decreasing = TRUE), ]

# Item fit — any misfitting items?
ifit <- item_fit(fit_alg)
ifit

# Visualize
plot(fit_alg, type = "icc")
plot(fit_alg, type = "wright")
plot(fit_alg, type = "info")
```

### Using Your Own Data

Your data should be a matrix or data frame with:

-   **Rows** = students (or examinees, respondents, etc.)
-   **Columns** = items (questions, tasks, etc.)
-   **Values** = 0 (incorrect) or 1 (correct)
-   **NA** = missing responses (handled automatically)

```{r}
# From a CSV file
my_data <- read.csv("my_test_data.csv", row.names = 1)
my_data <- as.matrix(my_data)

# Check it's formatted correctly
dim(my_data)        # (n_students, n_items)
table(my_data)      # should show only 0s and 1s
sum(is.na(my_data)) # count missing values

# Fit
fit_mine <- rasch_fit(my_data, seed = 123)
summary(fit_mine)
```

## Accessing the CmdStan Fit Object

For advanced analyses, you can access the underlying CmdStanMCMC object directly:

```{r}
# The raw CmdStan fit is stored in fit$fit
class(fit$fit)

# LOO cross-validation (requires the loo package)
# install.packages("loo")
# Note: requires log_lik in generated quantities (not yet in birt's Stan models)

# Extract raw draws for custom analyses
draws <- fit$fit$draws(variables = "beta", format = "draws_matrix")
dim(draws)  # (n_draws, n_items)

# Save a fitted model to disk
fit$fit$save_object("my_rasch_fit.rds")

# Load it back
# saved_fit <- readRDS("my_rasch_fit.rds")
```

## Choosing a Model

A practical guide to choosing between the three models:

1.  **Start with Rasch**. It's the simplest and most interpretable. If items fit well (outfit/infit between 0.7 and 1.3), stick with it.

2.  **Try 2PL** if Rasch item fit is poor or if you have theoretical reasons to believe items differ in discrimination. Compare the ICC plots — if 2PL curves have very different slopes, the discrimination parameter is meaningful.

3.  **Try 3PL** only if you have multiple-choice items and 500+ students. If the estimated guessing parameters are all near zero, drop back to 2PL.

## Troubleshooting

### Divergent Transitions

If you see "X divergent transitions", the sampler had trouble exploring the posterior. Try:

```{r}
fit <- rasch_fit(data, adapt_delta = 0.95, seed = 123)
```

The `adapt_delta` argument (passed through `...` to CmdStan) makes the sampler take smaller steps. Values between 0.9 and 0.99 usually help.

### Low ESS

If ESS values are below 400, try more iterations:

```{r}
fit <- rasch_fit(data, iter_sampling = 2000, seed = 123)
```

### Slow Compilation

The Stan model compiles to C++ the first time you run it. This takes 30-60 seconds but is cached afterward. Subsequent calls to the same model are fast.
